### 2022-06-13 paper.tex
shortened introduction

Sequencing-based assays with single-cell resolution have offered new means to understand the differences between the cells making up a sample. Single-cell RNA sequencing (scRNA-seq) techniques have matured at great pace in recent years. Now, methods to study epigenetics at single-cell resolution are rapidly catching up. Of special interest here is DNA methylation, i.e., the methylation of bases in the DNA, which is passed on to daughter cells in cell division. Especially in mammals, such methylation chiefly occurs at cytosine bases that are immediately after followed by guanine bases, so-called CpG \mbox{(5'--C--phosphate--G--3')} sites.

To detect CpG methylation by sequencing, bisulfite conversion is commonly employed: treatment of DNA with bisulfite causes unmethylated cytosine bases to be deaminated to uracil, which is replaced by thymine in subsequent PCR. Therefore, unmethylated CpGs are read as TG in sequencing of bisulfite-converted DNA while only methylated CpG sites are correctly read as CG \citep{Frommer_1992}. Such so-called bisulfite sequencing has been used for bulk samples since long, and methods for bisulfite sequencing with single-cell resolution were recently developed \citep{Smallwood_2014} and are increasingly used.

CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to spontaneous deamination), and the majority of existing CpG sites are methylated. Presence or absence of the methyl group at CpG sites is believed to affect gene regulation, with methylation inhibiting transcription. The removal of DNA methylation during cell differentiation is therefore conjectured to be an important mechanism to activate genome regions that are required in the target lineage and define cell identity \citep{cedar2022role}. One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes, for instance in embryonic development \citep{argelaguet2019gastru}.

In the present paper, we discuss strategies to analyze scBS data, suggest improvement to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.

@@

The standard approach to analyse scBS-Seq data is based on methodology developed for the analysis of single-cell RNA-Seq data. Therefore, we will start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analysed, before we discuss scBS data analysis.

The starting point in most scRNA-Seq analyses is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene. A first goal is then to assigning cell types or states to cells, and to this end, one needs to establish which cells are similar to each other, i.e., find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile. A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm. In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm. Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score. However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts. Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components. As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on very many genes. Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score. Hence, the PCA space representation y are suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

This procedure is commonly adapted when working with single-cell DNA-methylation data, so that established methods used in scRNA-Seq can be used.
The question is therefore how to construct a suitable matrix as input for the PCA. A simple and common approach, used for instance by \citet{luo2017single}, is to tile the genome into windows of, say, 100~kb size, and calculate for each cell the average methylation of the DNA within each window. To this end, we identify all CpG sites in the window for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell. Then we denote as average DNA methylation of this window in a given cell the proportion of observed CpG sites in the window that were found to be methylated (Figure \ref{fig:smoothres}A). This yields a matrix, with one row for each cell and one column for each window from the genome tiling, comprising numbers (methylation fractions) between 0 and 1. This matrix is now subjected to PCA.

We think that this analysis strategy has room for improvement, as we will discuss next. Here, we discuss the proposed improvements only qualitatively. For the mathematical details, please see the Methods section.

>>>>

Sequencing-based assays with single-cell resolution have offered new means to understand the differences between the cells making up a sample. Single-cell RNA sequencing (sc-RNA-seq) techniques have matured at great pace in recent years, with well developed analysis methodology, and methods to study epigenetics at single-cell resolution are rapidly catching up.

Here, we discuss bisulfite sequencing (sc-BS-Seq), a method to study DNA methylation. In brief, DNA is treated with bisulfite, which converts unmethylated cytosines to uracils which are read as thymine in subsequent PCR, while methylated cytosines are protected from conversion. After sequencing, these conversions allow to determine the methylation status of all cytosines covered by reads \citep{Frommer_1992}. Bisulfite sequencing can also be performed with single-cell resolution \citep{Smallwood_2014} and even in parallel with sc-RNA-Seq \citep{Clark2018}.

In the present paper, we discuss strategies to analyze sc-BS-Seq data, suggest improvement to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.

@@

The standard approach to analyse scBS-Seq data is based on methodology developed for the analysis of single-cell RNA-Seq data. Therefore, we start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analysed, before we discuss scBS data analysis.

The starting point in most scRNA-Seq analyses is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene. A first goal is usually to assigning cell types or states to cells. To this end, one needs to establish which cells are similar to each other, i.e., quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile. A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm. In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm. Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score. However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts. Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components. As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on very many genes. Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score. Hence, the PCA space representation is suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

This procedure is commonly adapted when working with single-cell DNA-methylation data, so that established methods used in scRNA-Seq can be used.
The task is therefore how to construct a suitable matrix as input for the PCA. A simple and common approach, used for instance by \citet{luo2017single}, is to tile the genome into windows of, say, 100~kb size, and calculate for each cell the average methylation of the DNA within each window. To this end, we identify all CpG sites in the window for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell. Then we denote as average DNA methylation of this window in a given cell the proportion of observed CpG sites in the window that were found to be methylated (Figure \ref{fig:smoothres}A). This yields a matrix, with one row for each cell and one column for each window from the genome tiling, comprising numbers (methylation fractions) between 0 and 1. This matrix is now subjected to PCA. After the PCA, one can proceed using approaches known from sc-RNA-Seq.

While this simple procedure is straight-forward and produces satisfactory results, there is room for improvement. Here, we propose three improvements that, as we show afterwards, considerably increase the information content of the processed data. We will first explain the proposed improvement in a qualitative manner, while mathematical details are given later in the Methods section. Then, we will demonstrate the improvement using benchmarks.
